---
title: "Week 2 homework"
output: html_document
date: '2022-03-08'
---

```{r setup, include=FALSE} 
knitr::opts_chunk$set(echo = TRUE)
```

## Goals. 
- Read `csv` files using `read_csv`
- Join tibbles/dataframes
- Use `mutate` to change values in the table, use `filter` to select rows that with match the criteria


### In the following code chunk, read "CD4.csv" as tibble "CD4" and "viral_load.csv" as tibble "vl" in the "homework" folder. Make sure load package `tidyverse` first. 
```{r}
library(tidyverse)
cd4 <- read_csv("CD4.csv")
vl <- read_csv("viral_load.csv")
```

### In the following code chunk, join the two tables that you generated from the two .csv files by matching 'study_id'. The joined_table will only have participants with both 'CD4' value and 'viral_load' value.  Which join function should be used? `full_join`, `left_join`, `right_join` or `inner_join`? Put the number of rows. 

```{r}
joined_table <- inner_join(cd4, vl, by='study_id')
```
### calculate the mean CD4 and viral load in the joined_table using funciton `mean()`, the viral load data will be presented as log10, and rounded to two decimals. Hint: use function `log10()` and `round()`.

```{r}
joined_table$CD4 %>% mean()
joined_table$viral_load %>% mean() %>% log10() %>% round(2)
```

### In the joined table, create a new column called "age_at_collect", calculated from "year_of_birth" and "year_of_collection".

```{r}
joined_table <- joined_table %>% mutate(
  age_at_collection = year_of_collection - year_of_birth
)
```

### Generate a new table called 'filtered_table', in which only males and age younger than 40 y/o at the time of collection will be stored. 

```{r}
filtred_table <- joined_table %>% filter(sex == 'M' & age_at_collection < 40)
```



