---
title: "Week 3 homework"
output: html_document
date: '2022-03-29'
---

```{r setup, include=FALSE} 
knitr::opts_chunk$set(echo = TRUE)
```

## Goals. 
- Use `tableby()` to summarize patient data table


### In the following code chunk, read "data.csv" as tibble "data". Make sure load package `tidyverse` first. 
```{r}
library(tidyverse)
data <- read_csv('data.csv')
```
### create a new column `age_group`, if the `age_at_collection` is greater or equal to 40 y/o, the value will be `≥ 40 y/o`, otherwise it will be `< 40 y/o`.

```{r}
data <- data %>% mutate(
  age_group = ifelse(age_at_collection >= 40, "≥ 40 y/o", "< 40 y/o")
)

```

### Use the `tableby()` function to summarize the `data` tibble. Group the patients by `sex`, summarize `CD4`, `viral_load`, `age_group`. Print out the table. Make sure you load library `arsenal` first. Use Kruskal-Wallis rank sum test (`kwt`) for CD4 value, one-way anova `viral_load`. Fisher's exact test for `age_group` (`fe`).
```{r, results='asis'}
library(arsenal)

tab <- tableby(sex ~ kwt(CD4) + viral_load + fe(age_group), data = data)
summary(tab)
```
